---
title: "COVID_case_estimation"
author: "Jonathan Moul"
date: "5/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document contains our attempt at an SIR model, together with the 'COVID_SIR.stan'
file. Note that to use this file, you need to manually alter the lines that set the directory and the filenames to read from (in the 'Preliminaries' and 'Data Processing' sections). 

## Preliminaries

```{r}
# Clear environment and set directory
rm(list=ls())
setwd("C:/Users/Jonathan/Documents/School Work/CS156b")

# Import data
county_totals = read.delim('nyt_us_counties.csv.txt', header=TRUE, sep=',', dec=".")
county_totals = county_totals[complete.cases(county_totals),]
county_pops = read.delim('county_populations.csv.txt', header=TRUE, sep=',', dec=".")
county_pops = county_pops[complete.cases(county_pops),]

# Load relevant libraries and ready parallel computing
library(rstan)
library(ggplot2)
library(bayesplot)
library(stringr)
library(rlist)
library(dplyr)
options(mc.cores = parallel::detectCores()) 
```

# Data Processing

We import the data. Then, we separate out counties with strict restaurant regulations over
the entire period from 4/5/20 to 5/10/20 and those with lax restaurant regulations over
the same period. We run our stan code on the summed case counts and population for each
group of counties and observe the differences between the posterior estimates of the
epidemiological parameters of the two groups.

```{r}
# Import the data
setwd("C:/Users/Jonathan/Documents/School Work/CS156b")
rests = read.delim('cdc_restaurant_orders2.txt', header=TRUE, sep=',', dec=".",
                   colClasses=c('character'))
rests = rests[complete.cases(rests),]
```

```{r}
# Process the data
# Combine the state and county FIPS
rests['fips'] = paste0(rests[,1], str_pad(rests[,2], 3, 'left', '0'))
# Change the classes of some columns
rests['date'] = as.Date(rests[,3], format='%m/%d/%Y')
rests['order_code'] = as.numeric(rests[,4])
```

```{r}
# Narrow our temporal window
now_rests = rests[rests['date'] > '2020-03-19' & rests['date'] < '2020-05-11']
dim(now_rests) = c(length(now_rests)/6, 6)

# Isolate the counties with harsh responses
harsh_fips = list()
lax_fips = list()
all_fips = unique(now_rests[,6])
for (curr_fips in all_fips) {
    # Get the table of only this fips
    curr_data = now_rests[now_rests[,6] == curr_fips,]
    dim(curr_data) = c(length(curr_data)/6, 6)
    colnames(curr_data) = c('fips_state', 'fips_county', 'date', 'code', 'state', 'fips')
    curr_data[,4] = as.numeric(curr_data[,4])
    if (min(curr_data[,4]) > 3) {
        harsh_fips = list.append(harsh_fips, curr_fips)
    }
    if (max(curr_data[,4]) < 4) {
        lax_fips = list.append(lax_fips, curr_fips)
    } 
}
```

```{r}
# Next, join the harsh_fips and lax_fips with their populations and cases
harsh_rests = now_rests[is.element(now_rests[,6], harsh_fips),]
colnames(harsh_rests) = colnames(rests)
lax_rests = now_rests[is.element(now_rests[,6], lax_fips),]
colnames(lax_rests) = colnames(rests)
harsh = merge(harsh_rests, county_totals, by.x=c('fips', 'date'), by.y=c('fips', 'date'))
harsh = merge(harsh, county_pops, by.x='fips', by.y='FIPS')
lax = merge(lax_rests, county_totals, by.x=c('fips', 'date'), by.y=c('fips', 'date'))
lax = merge(lax, county_pops, by.x='fips', by.y='FIPS')
```


```{r}
# Collate the harsh and lax tables by date
harsh = harsh[harsh['date'] > '2020-04-05' & harsh['date'] < '2020-05-11',]
lax = lax[lax['date'] > '2020-04-05' & lax['date'] < '2020-05-11',]

harsh_cases = harsh %>%
    group_by(date) %>%
    summarize(case_sum = sum(cases), pop_sum = sum(total_pop)) %>%
    pull(var=2)
harsh_pops = harsh %>%
    group_by(date) %>%
    summarize(case_sum = sum(cases), pop_sum = sum(total_pop)) %>%
    pull(var=3)
harsh_pop = max(harsh_pops)
lax_cases = lax %>%
    group_by(date) %>%
    summarize(case_sum = sum(cases), pop_sum = sum(total_pop)) %>%
    pull(var=2)
lax_pops = lax %>%
    group_by(date) %>%
    summarize(case_sum = sum(cases), pop_sum = sum(total_pop)) %>%
    pull(var=3)
lax_pop = max(lax_pops)
```

## Calculation and Analysis for Harsh Policy Group

```{r}
# Use stan on harsh_cases
N = length(harsh_cases)

# Input the data to stan
harsh_fit = stan('COVID_SIR.stan', iter = 25000, chains = 4,
           data = list(N=N, R_init=0, cases=harsh_cases, pop=harsh_pop))

# Print fitted model
print(harsh_fit)
```

We observe that the convergence behavior is very poor. This indicates that we should
likely try a variation of the model that needs to fit fewer parameters.

```{r}
# Get the results of the fit
harsh_fit_results = extract(harsh_fit)

# Record the mean sampled values of each parameter
harsh_I_init = mean(harsh_fit_results[['I_init']])
harsh_S_init = mean(harsh_fit_results[['S_init']])
harsh_gamma = mean(harsh_fit_results[['gamma']])
harsh_beta = mean(harsh_fit_results[['beta']])
harsh_I_vals = colMeans(harsh_fit_results[['I_vals']])
harsh_S_vals = colMeans(harsh_fit_results[['S_vals']])
harsh_R_vals = colMeans(harsh_fit_results[['R_vals']])
harsh_case_vals = colMeans(harsh_fit_results[['case_vals']])

# Plot the mean posterior values against the actual case numbers
x_vals = 1:N
pts = data.frame(x_vals, harsh_case_vals, harsh_cases)
ggplot(pts, aes(x=x_vals)) + 
    geom_line(aes(y=harsh_case_vals), color='blue') +
    geom_line(aes(y=harsh_cases), color='red') +
    ggtitle('Harsh Mean Posterior Fitted Cases vs Actual Cases') +
    xlab('Day from 4/10/20') +
    ylab('Number of People') +
    theme(legend.position = 'upper right')
```

```{r}
# Plot R_0 over time
x_vals = 1:N
harsh_r_0 = harsh_beta*harsh_S_vals/harsh_gamma
pts = data.frame(x_vals, harsh_r_0)
ggplot(pts, aes(x=x_vals)) + 
    geom_line(aes(y=harsh_r_0), color='blue') +
    ggtitle('Harsh R_0 Over Time') +
    xlab('Day from 4/10/20') +
    ylab('Est R_0') +
    theme(legend.position = 'upper right')
```

## Calculation and Analysis for Lax Policy Group

```{r}
# Use stan on lax_cases
N = length(lax_cases)

# Input the data to stan
lax_fit = stan('COVID_SIR.stan', iter = 25000, chains = 4,
           data = list(N=N, R_init=0, cases=lax_cases, pop=lax_pop))

# Print fitted model
print(lax_fit)
```

Once again, the convergence behavior is very poor, indicating that we should
likely try a variation of the model that needs to fit fewer parameters.

```{r}
# Get the results of the fit
lax_fit_results = extract(lax_fit)

# Record the mean sampled values of each parameter
lax_I_init = mean(lax_fit_results[['I_init']])
lax_S_init = mean(lax_fit_results[['S_init']])
lax_gamma = mean(lax_fit_results[['gamma']])
lax_beta = mean(lax_fit_results[['beta']])
lax_I_vals = colMeans(lax_fit_results[['I_vals']])
lax_S_vals = colMeans(lax_fit_results[['S_vals']])
lax_R_vals = colMeans(lax_fit_results[['R_vals']])
lax_case_vals = colMeans(lax_fit_results[['case_vals']])

# Plot the mean posterior values against the actual case numbers
x_vals = 1:N
pts = data.frame(x_vals, lax_case_vals, lax_cases)
ggplot(pts, aes(x=x_vals)) + 
    geom_line(aes(y=lax_case_vals), color='blue') +
    geom_line(aes(y=lax_cases), color='red') +
    ggtitle('Lax Mean Posterior Fitted Cases vs Actual Cases') +
    xlab('Day from 4/10/20') +
    ylab('Number of People') +
    theme(legend.position = 'upper right')
```

```{r}
# Plot R_0 over time
x_vals = 1:N
lax_r_0 = lax_beta*lax_S_vals/lax_gamma
pts = data.frame(x_vals, lax_r_0)
ggplot(pts, aes(x=x_vals)) + 
    geom_line(aes(y=lax_r_0), color='blue') +
    ggtitle('Lax R_0 Over Time') +
    xlab('Day from 4/10/20') +
    ylab('Est R_0') +
    theme(legend.position = 'upper right')
```

## Conclusion

Although the poor convergence diagnostics and the fitted curves' lack of resemblance to
the data indicate that any conclusions from this model are extremely tenuous at best, the
order of magnitude difference in the calculated reproduction constants R_0, as well as
the decrease of R_0 well below the reproductive threshhold of 1.0 for the counties with
harsh policies, indicate that a more thoughtfully constructed model of this type may
hold promise.